---
title: "Independent Project: Diagnostics for 2-sample t-test"
author: "brouwern@gmail.com"
date: "December 9, 2016"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Updates to this file will be announced via email and on teh course Facebook page
https://www.facebook.com/groups/930301587096169/

# Lion data set as 2-sample t-test

Example analysis of the lion data using a two-sample t-test.  


## Preliminaries
```{r}
#The following sets up the data fzo the analysis
#Set working directory

setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/Lecture/Unit3_regression/last_week")
```

Load data
```{r}
dat <- read.csv("lion_age_by_pop_and_sex.csv")
```




## Plot Lion data
```{r}
par(mfrow = c(1,1))
plot(age.years ~ portion.black, data = dat,
     main = "Lion pigment data split into 3 groups",
     ylim = c(0,19))
```

<br>

## Split lion data by groups

Make two groups, low pigmentation and high pigmentation, by converting the numeric pigmentation variable.

```{r}
par(mfrow = c(1,1))
pigment.mean <- mean(dat$portion.black)

plot(age.years ~ portion.black, data = dat,
     main = "Lion pigment data split into 2 groups",
     ylim = c(0,19))
abline(v = pigment.mean, col = 2, lwd = 3)

text(x = 0.125, y = 17, "Low\npigment", cex =1.1)

text(x = 0.895, y = 17, "High\npigment", cex = 1.1)
```


<br>

## Create Two pigment groups
```{r}
pigment.mean <- mean(dat$portion.black)

dat$pigment.groups.2 <- "high.pigment"

dat$pigment.groups.2[which(dat$portion.black <= pigment.mean)] <- "low.pigment"



dat$pigment.groups.2 <- factor(dat$pigment.groups.2)
summary(dat)


                               
dat$pigment.groups.2 <- factor(dat$pigment.groups.2,
                               levels =c("low.pigment",
                                          "high.pigment"))


table(dat$pigment.groups.2, dat$population)    
table(dat$pigment.groups.2, dat$sex)


```

**NOTE:** Splitting up a continous variable like this into a categorical variable and doing an ANOVA is **generally a BAD idea**. Regression is the proper way to analyze the raw data.  I am making these groups for illustration purposes only!

<br><br>

## Plot raw data by pigmentation group


```{r}
par(mfrow = c(1,1))
plot(age.years ~ pigment.groups.2,data = dat,
     xlab = "Pigmentation group",
     ylab = "Known Lion Age")
```

**Figure 1a:** Boxplots of lion ages in two pigmentation groups from the Serengeti and Ngorogoro Crater, Tanzania, east Africa.


## Plot means


```{r}
library(sciplot)


ci.fun. <- function(x) c(mean(x)-2*se(x), mean(x)+3*se(x))
lineplot.CI(x.factor = pigment.groups.2,
            response = age.years,
            data = dat,
            ci.fun = ci.fun.
              )
```

**Figure 2a:** Mean age of lions in two pigmentation groups from the Serengeti and Ngorogoro crater, Tanzania, east Africa.  Error bars are approximate 95% confidence intervals.



## 2-sample t-test

### The Normal way: 2-sample t-test w/ t.test

Test whether the two pigmentation groups have sig. different mean ages.
```{r}
t.test(age.years ~ pigment.groups.2, data = dat)
```


### The alternative way: t-test with lm()

You can set up basically the same test/model using the function lm() (though note that Welch's correction for unequal variances won't be applied).  Doing this is a useful way to look at the residuals of the t-test.

#### Fit lm() model
```{r}
lm.t.test <- lm(age.years ~ pigment.groups.2, data = dat)
```

#### Look at diagnostics of the t-test

Get the residuals
```{r}
resids <- resid(lm.t.test)
```


Plot diagnostic plots
```{r}
par(mfrow = c(2,2))
hist(resids)
plot(lm.t.test, which = 2)
plot(lm.t.test, which = 1)
plot(lm.t.test, which = 5)
```

* Normal QQ plot is a little off
* Residual's vs. fitted indicates some difference in variance between the groups


#### Fit lm() model to logged data

Check if log transformation improves thigns
```{r}
lm.t.test.log <- lm(log(age.years) ~ pigment.groups.2, data = dat)
```

#### Look at diagnostics of the t-test

Get the residuals
```{r}
resids.log <- resid(lm.t.test.log)
```


Plot diagnostic plots
```{r}
par(mfrow = c(2,2))
hist(resids.log)
plot(lm.t.test.log, which = 2)
plot(lm.t.test.log, which = 1)
plot(lm.t.test.log, which = 5)
```


* QQ is better
* Residvuals vs. fit is perhaps a little better.