---
title: 'Lecture 28: Multiple regression w/Lion data'
author: "brouwern@gmail.com"
date: "December 5, 2016"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/Lecture/Unit3_regression/last_week")
```

#Multiple linear regression

* Also known as "multiple regression"



## Lion data: predicting age from nose coloration

### Preliminaries

Load the data
```{r}
lions <- read.csv("lion_age_by_pop.csv" )
```

<br>

There are 2 populations, The main Serengeti population and the Ngorogoro crater (sub?) population
```{r}
summary(lions)
```

<br>
In the book they consider _all_ of the data combined and ignore the two separate populations (which are operate geographically and genetically)

<br>
Does the relationship between age and nose-blackness work the same for both populations?

<br>

### Plot lion data w/ both populations combined

We will "pool" the data from both populations (that is, ignore population differences)

#### Run regression
```{r}
#build regression model
m.pooled <- lm(age.years ~ portion.black, data = lions)
```

#### Look at summary of regression

```{r}
summary(m.pooled)
```


#### Plot pooled data
```{r}

#plot
plot(age.years ~ portion.black, data = lions, 
     cex = 2, lwd = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))
abline(m.pooled, lwd = 3, col = 2)

#To Do: add equations for line

```



### Plot lion data seperated by population

```{r}
i.Serengeti <- which(lions$population == "Serengeti")
i.Ngorogoro <- which(lions$population == "Ngorogoro")

#plot Serengeti
plot(age.years ~ portion.black, 
     data = lions,
     subset = i.Serengeti, 
     cex = 2, lwd = 2,col = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))


#plot Serengeti
points(age.years ~ portion.black, 
     data = lions,
     subset = i.Ngorogoro, 
     cex = 2, lwd = 2,
     col = 3,pch = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))

#add legend
legend("topleft",
       legend = c("Serengeti","Ngorogoro"),
       col = c(2,3),pch = c(1,2),cex = 1.2,
       )
```


#### What model best describes the data: 1 line or 2?


##### Hypothesis 0: there is no relationship

The null hypothesis Ho is that there is no relationship between how black the nose is and age.  We can see that there is a relationship and will skip this Ho.

##### Hypothesis 1: the populations are the same

The data can be modeled with a single regression line.  Any difference between the populations due to random noise.

```{r hypothesis 1a, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions,
      col = population, 
      shape = population) +
  geom_point(size = 5) + theme_bw() +
  ggtitle("")
```




```{r hypothesis 1b, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions) + 
  geom_smooth(method = "lm", se = F, size = 3#,
              #aes(linetype = population)
              )+
  geom_point(size = 5, aes(
      col = population, 
      shape = population)) +
  ggtitle("Hypothesis 1: one line") +
  theme_bw()
```



##### Hypothesis 2: the populations are different

*The populations have different biological/ecological processes going on that make the relationship between age and nose blackness depend on the population
*We say there is an **INTERACTION** between population and portion.black
*stated another way: The relationship between nose blackness and age is different between the two populations
*Therefore, the slope of the lines are different
*(This include the possibility that there is no relationship in one population (slope = 0) and is a relationship in the other (slope >0) )



<br><br>


```{r hypothesis 1, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions) + 
  geom_smooth(method = "lm", se = F, size = 3,
              aes(linetype = population,
                  col = population)
              )+
  geom_point(size = 5, aes(
      col = population, 
      shape = population)) +
  ggtitle("Hypothesis 2: two lines") +
  theme_bw()
```



#### Run multiple regression model

*We indicate in **interaction** in R using the multiple symbol "*"
```{r hypothesis 2, echo=FALSE}

m.by.pop <- lm(age.years ~ portion.black*population,                 data = lions)


```


<br>

Look at multiple regression output
```{r}
summary(m.by.pop)
```



#### Test models 

Compare the model w/1 line vs. the model w/ 2 lines

```{r}
anova(m.pooled, # data combined (aka "pooled")
      m.by.pop) # data seperated by population
```




#### MOre stuff

```{r}

#no relationship
m.null <- lm(age.years ~ 1,data = lions)

#relationship is the same for both populations
m.pooled <- lm(age.years ~ portion.black,data = lions)


#slope is the same but intercepts are different
m.by.pop.add <- lm(age.years ~ portion.black + population,data = lions)


#slopes AND intercepts are different
m.by.pop.intnx <- lm(age.years ~ portion.black*population,data = lions)

```



```{r}
#relationship is the same for both populations
summary(m.pooled)
```




```{r}
#slope is the same but intercepts are different
summary(m.by.pop.add)
```



```{r}
#slopes AND intercepts are different
summary(m.by.pop.intnx)
```

