---
title: "Lecture 29b: (Very) brief into to 2-way ANOVA"
author: "brouwern@gmail.com"
date: "December 9, 2016"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lion data set as 2-way ANOVA


```{r}
#The following sets up the data fro the analysis
#Set working directory

setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/Lecture/Unit3_regression/last_week")
```

Load data
```{r}
dat <- read.csv("lion_age_by_pop_and_sex.csv")
```




## Lion data
```{r}

plot(age.years ~ portion.black, data = dat,
     main = "Lion pigment data split into 3 groups",
     ylim = c(0,19))
```


## Split lion data by groups
```{r}
par(mfrow = c(1,1))
pigment.mean <- mean(dat$portion.black)

plot(age.years ~ portion.black, data = dat,
     main = "Lion pigment data split into 3 groups",
     ylim = c(0,19))
abline(v = pigment.mean, col = 2, lwd = 3)

text(x = 0.125, y = 17, "Low\npigment", cex =1.1)

text(x = 0.895, y = 17, "High\npigment", cex = 1.1)
```


<br>

## Create Two pigment groups
```{r}
pigment.mean <- mean(dat$portion.black)

dat$pigment.groups.2 <- "high.pigment"

dat$pigment.groups.2[which(dat$portion.black <= pigment.mean)] <- "low.pigment"



dat$pigment.groups.2 <- factor(dat$pigment.groups.2)
summary(dat)


                               
dat$pigment.groups.2 <- factor(dat$pigment.groups.2,
                               levels =c("low.pigment",
                                          "high.pigment"))


table(dat$pigment.groups.2, dat$population)    
table(dat$pigment.groups.2, dat$sex)


```

**NOTE:** Splitting up a continuous variable like this into a categorical variable and doing an ANOVA is generally a BAD idea. Regression is the proper way to analyze the raw data.  I am making these groups for illustration purposes only!

<br>

## Plot raw data by pigmentation group


```{r}
par(mfrow = c(1,1))
plot(age.years ~ pigment.groups.2,data = dat,
     xlab = "Pigmentation group",
     ylab = "Known Lion Age")
```

**Figure 1a:** Boxplots of lion ages in two pigmentation groups from the Serengeti and Ngorogoro Crater, Tanzania, east Africa.


## Plot raw data by pigmentation group


```{r}
par(mfrow = c(1,1))
plot(age.years ~ population,data = dat,
     xlab = "Pigmentation group",
     ylab = "Known Lion Age")
```

**Figure 1b:** Boxplots of lion ages in two populations the Serengeti and Ngorogoro Crater, Tanzania, east Africa.


<br><br><br>

## Plot means


```{r}
library(sciplot)


ci.fun. <- function(x) c(mean(x)-2*se(x), mean(x)+3*se(x))
lineplot.CI(x.factor = pigment.groups.2,
            response = age.years,
            data = dat,
            ci.fun = ci.fun.
              )
```

**Figure 2a:** Mean age of lions in two pigmentation groups from the Serengeti and Ngorogoro crater, Tanzania, east Africa.  Error bars are approximate 95% confidence intervals.



```{r}
lineplot.CI(x.factor = population,
            response = age.years,
            data = dat,
            ci.fun = ci.fun.
              )
```
**Figure 2b:** Mean age of lions in two population groups, the Serengeti and Ngorogoro crater, Tanzania, east Africa.  Error bars are approximate 95% confidence intervals.


## Calcualte the means of each of the FOUR groups

```{r}

source("plot_means.R")
#combine group names
dat$all.four.groups <- paste(dat$population,
                      dat$pigment.groups.2)

means <- tapply(dat$age.years,
       dat$all.four.groups,
       mean)

sds <- tapply(dat$age.years,
       dat$all.four.groups,
       sd)

ns <- tapply(dat$age.years,
       dat$all.four.groups,
       length)

summary(factor(dat$all.four.groups))

groups <- c("Ngorogoro high",
            "Ngorogoro low",
            "Serengeti high",
            "Serengeti low")
ses <- sds/sqrt(ns)
plot.means(means = means,
           SEs = ses,
           categories = groups,y.axis.label = "Lion Age")



```




<br><br>

## Two-way ANOVA

(again, don't do an analysis like this - use the original data and do an ANOVA!)

### Null model: age.years ~ 1

Age does not vary with pigmentation OR pigmentation
```{r}
m.null <- lm(age.years ~ 1 ,data = dat)
```

<br>

### Alternative model 1: age.years ~ pigment.groups.2
Hypoth:Age *does*  vary with pigmentation
```{r}
m.alt.1.pigment <- lm(age.years ~ pigment.groups.2,data = dat)
```


### Alternative model 2: age.years ~ population
Hypoth:Age varies with population
```{r}
m.alt.2.pop <- lm(age.years ~ population,data = dat)
```



### Alternative model 3: age.years ~ pigmentation + population
Hypoth:Age varies with population AND pigmentation

```{r}
m.alt.3.both <- lm(age.years ~ population +pigment.groups.2,data = dat)
```

<br>

### Alternative model 4: age.years ~ pigmentation *population (FUll model)

Hypoth:Age varies with population
```{r}
m.alt.4.intxn <- lm(age.years ~ population*pigment.groups.2,data = dat)

m.alt.4.mean <- lm(age.years ~ -1+ population:pigment.groups.2,data = dat)
```

<br><br>

### THe hard way: Omnibus ANOVA F-test on EACH set of models 


Null vs. alt.1 (pigmentation groups)
```{r}
anova(m.null, m.alt.1.pigment)
```

<br><br>

Null vs. alt.3 (populations)
```{r}
anova(m.alt.1.pigment, m.alt.2.pop)
```



Null vs. alt.3 (pigmentation + population)
```{r}
anova(m.alt.1.pigment, m.alt.3.both)
```



Null vs. alt.4 (pigmentation*population interaction)
```{r}
anova(m.alt.3.both, m.alt.4.intxn)
```



### The easy way


anova() on "interaction" model
```{r}
anova(m.alt.4.intxn)
```






### Writing up result of Omnibus F-test

*Summarize reulst of ANOVA in a single sentence:*
"There was a significant difference in the mean ages of the three diffrent pigmentation groups (F = 106.67, p < 0.00001, DF = 92,90)"

<br><br>

## Comparisons using Tukey-HSD

### Refit m.alt w/ aov()

```{r}

dat$pigment.groups.2.rename <- gsub("pigment","",dat$pigment.groups.2)

m.alt.4.inxtn.aov <- aov(age.years ~ pigment.groups.2.rename*population,data = dat)
```

<br><br>

### Pairwise comparison w/TukeyHSD()

Run TukeyHSD()
```{r}
my.tukey.intxn <- TukeyHSD(m.alt.4.inxtn.aov)
```

<br><br>

