---
title: "Multiple regression"
author: "brouwern@gmail.com"
date: "December 5, 2016"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Sample Paper Using Lion data

A an example of how to write and format your final project, using lion data from Whitlock & Shulter chapte 17, which is originally from Whitmas et al 2004 in Science.


# Preliminaries


```{r}
#The following sets up the data fro the analysis
#Set working directory

setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/Lecture/Unit3_regression/last_week")
```

Load data
```{r}
dat <- read.csv("lion_age_by_pop.csv")
```

# Content of a scientific paper

<br>


## Does does relationship between nose pigmentation and age vary between different populations of African lion (_Panthera leo_)?

## ABSTRACT

For information on the format of "Structured abstacts", see
https://www.nlm.nih.gov/bsd/policy/structured_abstracts.html

**INTRODUCTION**: Being able to accurately understand the age structure of population
**OBJECTIVES**: The primary of objective of this study was to determine if the age of lions can be predicted from the amount of pigementation on their noses, and whether this relationship varies between populations.
**METHODS**:(3-4 sentneces on methods; likely to be longest or 2nd longest part of abstract)
**RESULTS**:(2-5 sentences on results; likely to be longest or 2nd longest part of abstract)
**CONCLUSION**(2-3 sentences stating the biological/ecological/scientific conclusions that can be drawn from the study).

## INTRODUCTION

*Please include 3 to 4 citations from **peer reviewed journal articles** here.
*(There needs to be 4 references in the papers; at least 3 have to be in the introduction)
*References should be properly formated, ie "Brouwer et al (2015)."
*Use your introductory material from the previous assignment
*Please end your intro with a sentence or two stating what the overall goal/hypothesis /question of the study is.
*Please update your intro based on the suggestions
*I do not expect you to write a perfect introduction, but I expect to see improvement from your first draft

<br>

## METHODS

### _Study species_
*If you study is focused on a particular species include a few sentences about its basic biology in case your reader is not an expert on that species.
*All statements of fact should be cited
*ig, "The African lion (_Panthera leo_) is the largest carnivore in the world and can live to ages of ... (citation).  While endangered in much of its range, trophy hunting is legal in parts of Africa (cite) ...""

### _Study site_
*If your study occurred at a particular location that someone would have to re-visit if they wanted to replicate or follow up on your, provide information about where it is located (State, County, etc) and what the physical conditions are like.
*ie, "This study was conduct on the Serengeti of Tanzania and teh Ngorogoro crater.  The Serengeti is ... (cite).  Ngorogo crater is ... (cite)."

## _Field/Lab methods_
*How data was collected

## _Statistical methods_
I tested whether there was a significant relationship between nose pigmentation and lion age using linear regression.  To determine if this relationship varied between lion populations I tested for an effect of population and a  pigmenation*popualtion interaction.  All analyses were carried out in R 3.3.1 (R Core Team 2016).

## RESULTS

### _Descriptive statistics_
*You should start you results with a few sentences about the general characteristics of the data you collected

Data on nose pigmentation and age was collected for 32 lions; 22 lions were from the Serengeti population and 10 from Ngorogoro.  (Raw data are availabe in Table A1 in Appendix 1).  The mean age of lions in The Serengeti popualtion was   The distribution of ages for each population is shown in Figure A1.

```{r,, include=FALSE}
dim(dat)
summary(dat)

lion.means <- tapply(dat$age.years,
      dat$population,
      FUN = mean)

lion.sd <- tapply(dat$age.years,
      dat$population,
      FUN = sd)

lion.n <- tapply(dat$age.years,
      dat$population,
      FUN = length)

lion.se <- lion.sd/sqrt(lion.n)

```


## DISCUSSION

*A discussion section is optional
*Up to 5 points will be given as extra credit for a reasonably well-written discussion.  THe discussion should include references

## ACKNOWLEDGEMETNS

*Recognize anyone who helped you on the project, such as professor who you consulted with on the project, roommates who helped you collect data, or classmates who helped you with stats or code.

## REFERENCES

*You should have 4 references from peer reviewed journals
*Your 5th reference should be for R, as shown below.

R Core Team (2016). R: A language and environment for statistical computing. R Foundation for
Statistical Computing, Vienna, Austria. https://www.R-project.org/.

## FIGURES

**Figure 1:**

## TABLES

## APPENDIX 1: Raw data table

## APPENDIX 2

## APPENDIX 3




#Multiple linear regression

* Also known as "multiple regression"



## Lion data: predicting age from nose coloration

### Preliminaries

Load the data
```{r}
lions <- read.csv("lion_age_by_pop.csv" )
```

<br>

There are 2 populations, The main Serengeti population and the Ngorogoro crater (sub?) population
```{r}
summary(lions)
```

<br>
In the book they consider _all_ of the data combined and ignore the two seperate populations (which are sperate geographically and genetically)

<br>
Does the relationship between age and nose-blackness work the same for both populations?

<br>

### Plot lion data w/ both populations combined

We will "pool" the data from both populations (that is, ignore population differences)

#### Run regression
```{r}
#build regression model
m.pooled <- lm(age.years ~ portion.black, data = lions)
```

#### Look at summary of regression

```{r}
summary(m.pooled)
```


#### Plot pooled data
```{r}

#plot
plot(age.years ~ portion.black, data = lions, 
     cex = 2, lwd = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))
abline(m.pooled, lwd = 3, col = 2)

#To Do: add equations for line

```



### Plot lion data seperated by population

```{r}
i.Serengeti <- which(lions$population == "Serengeti")
i.Ngorogoro <- which(lions$population == "Ngorogoro")

#plot Serengeti
plot(age.years ~ portion.black, 
     data = lions,
     subset = i.Serengeti, 
     cex = 2, lwd = 2,col = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))


#plot Serengeti
points(age.years ~ portion.black, 
     data = lions,
     subset = i.Ngorogoro, 
     cex = 2, lwd = 2,
     col = 3,pch = 2,
     xlim = c(0.0, 0.9),
     ylim = c(0,14))

#add legend
legend("topleft",
       legend = c("Serengeti","Ngorogoro"),
       col = c(2,3),pch = c(1,2),cex = 1.2,
       )
```


#### What model best describes the data: 1 line or 2?


##### Hypothesis 0: there is no relationship

THe null hypothesis Ho is that there is no relationship between how black the nose is and age.  We can seee that there is a relationshop and will skip this Ho.

##### Hypothesis 1: the populations are the same

The data can be modeled with a single regression line.  Any differene between the popuatliosn due to random noise.

```{r hypothesis 1a, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions,
      col = population, 
      shape = population) +
  geom_point(size = 5) + theme_bw() +
  ggtitle("")
```




```{r hypothesis 1b, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions) + 
  geom_smooth(method = "lm", se = F, size = 3#,
              #aes(linetype = population)
              )+
  geom_point(size = 5, aes(
      col = population, 
      shape = population)) +
  ggtitle("Hypothesis 1: one line") +
  theme_bw()
```



##### Hypothesis 2: the populations are different

*The populations have different biological/ecological processes going on that make the relationship between age and nose blackness depend on the population
*We say there is an **INTERACTION** between population and portion.black
*stated another way: The relationship between nose blackness and age is different between the two populations
*Therefore, the slope of the lines are different
*(THis include the possiblity that there is no relationship in one population (slope = 0) and is a relationship in the other (slope >0) )



<br><br>


```{r hypothesis 1, echo=FALSE}
library(ggplot2)

qplot(y = age.years,
      x = portion.black, 
      data = lions) + 
  geom_smooth(method = "lm", se = F, size = 3,
              aes(linetype = population,
                  col = population)
              )+
  geom_point(size = 5, aes(
      col = population, 
      shape = population)) +
  ggtitle("Hypothesis 2: two lines") +
  theme_bw()
```



#### Run multiple regression model

*We indicate in **interaction** in R using the multiple symbol "*"
```{r hypothesis 2, echo=FALSE}

m.by.pop <- lm(age.years ~ portion.black*population,                 data = lions)


```


<br>

Look at multiple regression output
```{r}
summary(m.by.pop)
```



#### Test models 

Compare the model w/1 line vs. the model w/ 2 lines

```{r}
anova(m.pooled, # data combined (aka "pooled")
      m.by.pop) # data seperated by population
```




#### MOre stuff

```{r}

#no relationship
m.null <- lm(age.years ~ 1,data = lions)

#relationship is the same for both populations
m.pooled <- lm(age.years ~ portion.black,data = lions)


#slope is the same but intercepts are different
m.by.pop.add <- lm(age.years ~ portion.black + population,data = lions)


#slopes AND intercepts are different
m.by.pop.intnx <- lm(age.years ~ portion.black*population,data = lions)

```



```{r}
#relationship is the same for both populations
summary(m.pooled)
```




```{r}
#slope is the same but intercepts are different
summary(m.by.pop.add)
```



```{r}
#slopes AND intercepts are different
summary(m.by.pop.intnx)
```

